---
title: "Visualization with ggplot2"
author: "Irfan Kanat"
date: "11/04/2015"
output: 
  pdf_document:
    fig_width: 4
    fig_height: 3
    fig_caption: false
---


R has long been known for its extensive visualization capabilities. The number of packages that handle visualizations are many, yet ggplot shines among them all. Today I will focus on ggplot and discuss plotting histograms and scatter plots with qplot. I will focus mostly on qplot() function, and discuss ggplot structure only briefly.

## Introducing the Dataset

[In this document we will analyze the Motor Trends data.](http://www.jstor.org/stable/2530428) The dataset was compiled from 1974 issues of Motor Trends magazine and is included with R Base package.
  
Let us start with loading the dataset.
  
```{r}
data(mtcars)
```

As we learned in the section on packages, you can querry the documentation for almost anything. Including the datasets included in packages. The document includes descriptions of the variables.

```{r}
?mtcars
```

Let us get a sense of the data.

```{r}
# A summary of variables
summary(mtcars)
# Correlation table for first 4 variables (due to space concerns)
cor(mtcars[,1:4])
# bivariate comparisons of categorical variables
table(mtcars[,c("am","cyl")])

# The histogram below should reflect these figures. 
table(mtcars$gear)
```

## Plotting with qplot()

Now we can get to the fun part. qplot simplifies the ggplot functionality by automating most common tasks. We will use qplot for most common plots. 

```{r}
# Load the ggplot package
library(ggplot2)
# Review function syntax
?qplot
```

### Histogram

You would use a histogram when you are interested in frequencies of certain categories, like number of people with different eye colors.

```{r}
# Let us report the number of cars with differing number of front gears
qplot(factor(gear), data=mtcars, geom="bar") # used factor to declare categorical
```

If we want to get fancy and want to report across two categorical variables we can color the bars based on another variable.

```{r}
qplot(factor(gear), data=mtcars,  fill=factor(am), geom="bar") # used factor to declare categorical
```

### Scatter Plots

If you are interested in the relationship between two continuous variables, you can use scatter plots. 

```{r}
qplot(hp, mpg, data=mtcars)
```

Let us impose an additional factor into the plot. Let us color the dots by the number of cylinders.

```{r}
qplot(hp, mpg, data=mtcars, color=factor(cyl))
```

Size of dots dependent on a continuous variable (displacement).

```{r}
qplot(hp, mpg, data=mtcars, color=factor(cyl), size=disp, alpha=.08)
```

Let us fit a regression line. This is where things start to get a bit ggplotty.

```{r}
qplot(hp, mpg, data=mtcars) + 
  geom_smooth(method=lm, sd=F)
```

## ggplot

qplot provides a convenient command for plotting. While qplot would address 90% of your plotting needs. ggplot is way more than qplot, it is almost a different language just for plotting. The intricacies may be hard to learn and is clearly beyond the scope of this workshop. I am providing ggplot code below to achieve the same results as the qplot, so the attendees can get a sense of what ggplot is really about.

### Histogram

```{r}
# Initialize the plot with variables of interest
ggplot(mtcars, aes(factor(gear))) +  
# Instruct ggplot to plot bars of width .3
  geom_bar(stat = "bin", width=0.3) +
  ggtitle('GGPlot is Highly Customizable') +
  xlab('Forward Gears') +
  ylab('Nr of Cars') 
```

### Scatter Plot

```{r}
ggplot(mtcars, aes(x=hp, y=mpg)) +  
  geom_point(aes(color=factor(cyl), size=disp)) + # For scatter plot
	geom_smooth(method=lm) + # Add a regression line
  ggtitle('GGPlot FTW!') # Add a title
```

### Bar Charts 

You use bar charts when you want to visualize the relationship of a continuous variable over a categorical variable (eg. gender-height). Here I plot mean mpg over two categorical variables. 

```{r}
ggplot(mtcars,aes(x=factor(gear),y=mpg,fill=factor(vs)), color=factor(vs)) +  
  stat_summary(fun.y=mean, position=position_dodge(), geom="bar")
```

### Bonus! World Map Visualizations

After having mingled that GDP data so much, I could not help but plot the results on a map. Here is how:

Process data into a map.

```{r}
# Load the Library
library(rworldmap) # Install if necessary

# Get the dataset
cData_Long<-read.csv("cDataLong.csv")

# Subset
cDataL2011 <- subset(cData_Long, time==2011)

# Get a taste
head(cDataL2011)

# Turn into map
cDataL2011<- joinCountryData2Map(cDataL2011, joinCode = "ISO2",
                                 nameJoinColumn = 'ISO2')
```

Plot the world map.

```{r, fig.width=6, fig.height=4}
mapCountryData(cDataL2011, nameColumnToPlot = "GDP")
```

Focus on Eurasia.

```{r, fig.width=6, fig.height=4}
mapCountryData(cDataL2011, nameColumnToPlot = "GDP", mapRegion = "eurasia")
```

Plot a categorical variable as color and continuous as size.

```{r}
mapDevice() # Initialize the map
mapBubbles(dF=cDataL2011, nameZSize="GDP", nameZColour="Continent", colourPalette="rainbow", oceanCol="lightblue", landCol="wheat", alpha = .8)
```

## caret and Visualizations

We will cover caret a little in machine learning samples. This package provides convenient shortcuts to ggplot functionality. Simplifying most common plotting tasks in machine learning. Please [refer to the project page](http://topepo.github.io/caret/visualizations.html) for further reference.