---
title: "Assignment"
author: "Irfan Kanat"
date: "11/09/2015"
output: pdf_document
---

# R Workshop Assignment

Before coming to the second session, please make sure you have the packages listed below installed and running.

There may be some configuration problems with your installation, and you don't want to trouble shoot at the last minute. So act quickly on them and make sure you have a running environment before the workshop.


## Installing rJava

Some of the packages we use will require rJava to be installed. Unfortunately rJava is not available through CRAN. Follow the discussion [here to learn how to install rJava in Windows.](http://stackoverflow.com/a/7604469/3475252)

Basically you need to do:

1 - Add java jvm libraries to your R path.
2 - Restart R
3 - Install the rJava from R Forge

```{r, eval=FALSE}
install.packages('rJava',,'http://www.rforge.net/')
```

As I said, [look here for details.](http://stackoverflow.com/a/7604469/3475252) I can not provide you with specific instructions on this one as I don't use Windows. 


## Installing caret Package

I noticed some of you had difficulty installing caret package. Please try the code below. It should just work.

```{r, eval=FALSE}
install.packages("caret", dependencies = c("Depends", "Suggests"))
```

## Install Others

**partykit:** Despite the fun name, the partykit is a visualization package for tree based models.

**ggplot2:** This is the package we will use for visualization examples.

**dplyr:** This is the package we will use for manipulation examples.

**reshape:**

```{r, eval=F}
install.packages(c("partykit", "ggplot2", "dplyr", "reshape"))
```

